Staged Abstract Interpreters

### 2019-06-29 paper/10_intro.tex
revise intro

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close
connection between compilers and interpreters. The first Futamura projection
specifically shows that specializing an interpreter with respect to the input
program yields an equivalent executable. Partial evaluation
\cite{DBLP:books/daglib/0072559} was the first proposed approach to realize
Futamura projections: it identifies the binding-time of variables in
the program (i.e., they can be known statically or dynamically),
evaluates the static part, and finally generates a residual program that solely
relies on the dynamic part. However, given an arbitrary program, precisely
analyzing its binding-times is hard in general. As an alternative and pragmatic
approach to specialization and partial evaluation, multi-stage programming (MSP) 
\cite{taha1999multi, DBLP:conf/pepm/TahaS97} requires the staging
annotations (i.e., binding-time annotations) to be given explicitly from the
programmers. The staging annotations can be either syntactic (as in
MetaML/MetaOCaml, through quasi-quote, escape, etc.) or type-based \cite{DBLP:conf/gpce/RompfO10}.
These staging annotations identify which part of the inputs should be
specialized. As a classical example, we use the power function and type-based
annotations to introduce the idea of MSP:

\begin{lstlisting}
  def power(b: Rep[Int], e: Int): Rep[Int] = if (e == 0) 1 else b * power(b, e - 1)
\end{lstlisting}

If the programmer identifies that @b@ will be known in the future stage (as shown
on its type @Rep[Int]@ -- a representation of @Int@) and @e@ is known at the
current stage, say 5, then we can specialize the function @power@ with
respect to @e = 5@, and generate a specialized function where the overhead of
recursion is eliminated. Conceptually, the generated code looks like the following:

\begin{lstlisting}
  def power5(b: Int): Int = b * b * b * b * b
\end{lstlisting}

Now to concretely illustrate the idea of Futamura projection, consider that we
have an interpreter @eval@ of some language:

\begin{lstlisting}
  def eval(e: Expr)(arg: Input): Value = ...
\end{lstlisting}

where $e$ is a program of type @Expr@ and @arg@ is the input to $e$,
and @eval@ produces a value.
Semantically, the interpreter satisfies $ \texttt{eval}(e)(arg) = [\![ e ]\!] arg$ for all
programs and inputs. Given a program $e_0$, by applying the first Futamura
projection, we can obtain a specialized interpreter
$\texttt{eval}_{\texttt{e0}}$ with respect to $e_0$, which is the so-called \textit{equivalent executable}. 
By definition of the interpreter, 
$\texttt{eval}_{\texttt{e0}}(arg) = [\![ e_0 ]\!] arg $. If we generalize the
argument @arg@ to an environment that maps free variables in $e$ to some
values, the interpreter is a standard environment-passing interpreter,
where the specialization through MSP is also applicable.

Roughly at the same time when Futamura projections were envisioned in the 1970s,
\citet{DBLP:conf/popl/CousotC77} proposed abstract interpretation as a
lattice-based semantic approach to construct sound static analysis, by
approximation of fix-points. However, constructing sound abstract interpreters
was considered abstruse and complicated for a long time.
Recent progress such as Abstracting Abstract Machines (AAM)
\cite{DBLP:conf/icfp/HornM10} uncovers a systematic method to derive sound
abstract interpreters from their concrete counterparts, where the soundness can
be easily established by examining the transformation of semantics. The AAM
approach also has been applied to different variants of definitional
interpreters and abstract machines \cite{DBLP:journals/jfp/HornM12,
DBLP:conf/icfp/HornM10, DBLP:journals/pacmpl/DaraisLNH17}.

\vspace{-1em}
\begin{figure}[h]
  \begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=3em,minimum width=2em]
  {
    \begin{smallmatrix} \text{unstaged} \\ \text{abstract} \\ \text{interpreters} \\ (\text{Sec.}~\ref{unstaged_abs}) \end{smallmatrix} &
    \begin{smallmatrix} \text{staged} \\ \text{abstract} \\ \text{interpreters} \\ (\text{Sec.}~\ref{sai}) \end{smallmatrix} \\
    \begin{smallmatrix} \text{unstaged} \\ \text{concrete} \\ \text{interpreters} \\ (\text{Sec.}~\ref{unstaged_conc}) \end{smallmatrix} &
    \begin{smallmatrix} \text{staged} \\ \text{concrete} \\ \text{interpreters} \\ (\text{Sec.}~\ref{stagedinterp}) \end{smallmatrix} \\ };
  \path[-stealth]
    (m-1-1) edge node [above, font=\scriptsize] {$\updownarrows$} (m-1-2)
    (m-2-1) edge node [left, font=\scriptsize] {$\alpha$} (m-1-1)
    (m-2-1) edge node [below, font=\scriptsize] {$\updownarrows$} (m-2-2)
    (m-2-2) edge node [right, font=\scriptsize] {$\alpha$} (m-1-2);
  \end{tikzpicture}
  \caption{The confluence of specialization and approximation.}
  \label{confluence}
\end{figure}
\vspace{-1em}

\paragraph{Futamura Projection of Abstract Interpreters}

Now that we have seen two orthogonal transformations of interpreters: by
specialization, we can refactor interpreters to code generators; by
approximation, we can refactor interpreters to static analyzers. As an
intellectual quest, we should naturally ask -- starting from an abstract
interpreter, \textit{can we derive a sound abstract interpreter while it
produces optimized code that does the analysis?}
Conceptually, it is similar to the concrete setting: given an
abstract interpreter $\widehat{\texttt{eval}}: \texttt{Expr} \to
\widehat{\texttt{Env}} \to \widehat{\texttt{Value}}$, which takes a program, an abstract
environment and returns abstract values, we would like to specialize it with
respect to some program $e_0$ and produce a compiled analysis
$\widehat{\texttt{eval}}_{\texttt{e0}} : \widehat{\texttt{Env}} \to
\widehat{\texttt{Value}}$, where the abstract environment $\widehat{\rho}$ is kept as input.
The specialized analysis satisfies $\widehat{\texttt{eval}}_{\texttt{e0}}(\widehat{\rho}) =
\widehat{\texttt{eval}}(e_0)(\widehat{\rho})$.

In this paper, we study the confluence of two old ideas --- Futamura projection
and abstract interpretation, but from the perspective of their very recent
realizations --- multi-stage programming and definitional abstract interpreters.
To be specific, we present the application of the first Futamura projection on
definitional abstract interpreters, which intersects as staged abstract
interpreters. We develop a unified approach to fulfill the four different
semantics, as shown in Figure~\ref{confluence}. Starting with a generic but
parametric interpreter shared by the four semantics, by simply instantiating
the interpreter differently, we can automatically derive a concrete interpreter,
or an abstract interpreter, or their staged versions that generate code, respectively.

\iffalse
embedded domain-specific languages \cite{DBLP:conf/snapl/RompfBLSJAOSKDK15,
DBLP:journals/jfp/CaretteKS09, DBLP:conf/icfp/GibbonsW14,
Hofer:2008:PED:1449913.1449935},
The key idea that enables this flexibility is to use abstract type members first
abstract over concrete/abstract components (e.g., concrete values or abstract
values), then as well as the binding-time of them (e.g., static or dynamic). By
staging, the overhead caused by the monadic layers is eliminated in the
generated code.
\fi

\paragraph{Abstraction without Regret}

To enable such flexibility between different semantics, we combine generic
programming and generative programming, and extensively use abstract type
members over the following abstractions, which can be later instantiated.

\begin{itemize}
\item Env-store-value abstractions, which define the domains that the interpreter will be running with.
  The environments, stores, and values are initially declared as abstract types in the generic interpreter.
  As shown by the AAM approach \cite{DBLP:journals/jfp/HornM12,
  DBLP:conf/icfp/HornM10}, if we widen the values to abstract domains (such as
  intervals) and refactor the store to a mapping from finite addresses to sets of
  values, we can derive a computable and sound abstract interpreter.
\item Binding-time abstractions, which control whether the interpreter produces
  values or generates code. The binding-times differentiate whether data are 
  known statically or dynamically. Unsurprisingly, such binding-time
  information can be encoded into higher-kinded types 
    \cite{DBLP:journals/jfp/CaretteKS09, Ofenbeck:2017:SGP:3136040.3136060}.
\item Monadic abstractions, which encapsulate the above two abstractions.
  The interpreter itself is written in monadic form, and by plugging different monads, 
  the underlying semantics of the interpreter is changed. We also use monad transformers to combine multiple monadic effects.
  %\todo{mention that direct-style also works?}
  %\cite{Steele:1994:BIC:174675.178068, DBLP:conf/popl/LiangHJ95, DBLP:journals/pacmpl/DaraisLNH17, Sergey:2013:MAI:2491956.2491979}
\end{itemize}

We show that these well-known abstractions can be combined smoothly, and more
importantly for free. We implement the four interpreters using Scala and the
Lightweight Modular Staging \cite{DBLP:conf/gpce/RompfO10} framework. Through
staging, the major overhead of abstractions -- from the monadic layers -- is
completely eliminated in the generated code. We observe that the staged version is 
on average 11x times as fast as the unstaged abstract interpreter. 
However, we do not lose any advantages of this high-level programming style,
e.g., modularity and extensibility. The meaning of generated programs also remain
the same, particularly, the soundness of static analysis after staging.

%\paragraph{Performance via Specialization}
Besides the intellectual merit shown in the confluence diagram, this paper
contributes to a practical problem in static analysis: \textit{can we speed up a
static analyzer without any intrusive changes and soundness compromises?} Static
analysis is known as a trade-off game between \textit{precision} and
\textit{efficiency}, we argue that applying meta-programming (multi-stage
programming, in particular) with abstract interpreters is an effective approach
to improve the latter while the former is untouched.
In the literature, specializing analysis with respect to the input program
\cite{damian1999partial, amtoft1999partial, Boucher:1996:ACN:647473.727587,
ashley:practical} is not new, despite the fact that most of the previous works
are ad-hoc on a particular analysis or requiring significant changes to the
analyzer. For example, abstract compilation is one of the techniques
\cite{Boucher:1996:ACN:647473.727587}. But it requires the whole analyzer to be
rewritten to the closure generation form. In this paper, we advocate a systematic
approach of generative programming that achieves the performance goal with less
intrusive changes and engineering efforts.

\paragraph{Applications}

Furthermore, we show various applications of staging an abstract interpreter.
(1) We compare our approach with abstract compilation, and show that we can
achieve the same optimization by utilizing type-based stage annotations,
meanwhile, the analyzer program does not need any changes. (2) We extend the 
basic staged abstract interpreter to diverse control-flow analysis, 
including a store-widened analysis, a context-sensitive analysis and abstract garbage collection.
(3) Staging abstract interpreter enables modular analysis and provides a mechanized
way to create library summaries. We notice that we may apply the staged abstract
interpreter separately on the different part of the analyzed program, and the
generated code is reusable and composable. Therefore, we can turn a
whole-program analyzer into an analyzer that runs modularly.

\iffalse
Our next contribution is to enable modular analysis based on a whole-program
analyzer by meta-programming. Modern software are shipped with large libraries
\cite{toman_et_al:LIPIcs:2017:7121}, and analyzing programs with large libraries
is usually expensive or unnecessarily repeated for whole-program analyzing
algorithms. For example, it has been shown that analyzing a simple ``Hello
World'' program in Java depends on additional 3,000 classes in the library
\cite{DBLP:conf/oopsla/KulkarniMZN16}. A natural idea is to analyze libraries
and generate summaries, later we can plug in the necessary information into the
summaries and finish the analysis; therefore such summaries should be modular
and reusable. Meta-programming gives us a chance to readily tweak a
whole-program analyzing algorithm into an analyzer that works modularly and
produces summaries. Still, the idea is to specialize the abstract interpreter
that was used for whole-program analysis but on a subset of the program, for
examples, a set of generic functions for list manipulation. Such specialization
works on lambda expressions and generates code, which is partial analysis
results and can be completed by running with the necessary arguments (for
example, an environment that closes the lambda term) at next stages. Multiple
partial analysis results also can be reused later and composed together as they
are just normal programs at the next stage. Through staging, we mechanically
obtain a modular analysis, even though the original algorithm is formulated as a
whole-program analysis.
\fi

\paragraph{Contributions} Briefly, the contribution of the paper is twofold:
\begin{enumerate}[leftmargin=2em]
  \item Intellectually, through multi-stage programming, we present a unified construction 
    that naturally extends the first Futamura projection to abstract interpreters.
 \item Pragmatically, by case studies and empirical evaluation, we show that staging 
   abstract interpreters is useful to improve performance while not touching
   the soundness. 
\end{enumerate}
\paragraph{Organization} The paper is organized as follows:
\begin{itemize}[leftmargin=2em]
\item Starting from a generic interface of the interpreter for a small higher-order language
  (Section~\ref{prelim}), we first show the instantiation of concrete
  interpretation (Section~\ref{unstaged_conc}), and then lift it to the staged
  version by replacing the binding-time types (Section~\ref{stagedinterp}).
\item We present the unstaged abstract interpreter under the same framework by
  replacing the environment, store and values (Section~\ref{unstaged_abs}).
  Finally, we show the combination of approximation and specialization, dubbed
  \textit{staged abstract interpreters}, can be readily derived (Section~\ref{sai}).
\item We conduct three case studies on the applicability of our approach
  (Section~\ref{cases_study}).
  \begin{enumerate}
  \item We first revisit the Abstract Compilation (AC)
    \cite{Boucher:1996:ACN:647473.727587} (Section~\ref{cs_ac}), showing that we can
    achieve the same optimization with much less engineering effort.
  \item We extend the staged abstract interpreter to various control-flow
    analysis (Section \ref{cfa}), including context-sensitivity, store-widening, and abstract
    garbage collection, showing that our approach is widely applicable.
  \item We apply our approach to modular analysis (Section \ref{modular}),
    i.e., analyzing the components of the program separately and composing them later,
    showing a novel application of generative programming with program analysis.
  \end{enumerate}
\item We empirically evaluate the performance improvement by staging on
    control-flow analysis (Section~\ref{evaluation}). We compare the both
    context-insensitive and store-widening analysis.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\iffalse
Likewise, specializing static analyses by partial evaluation emerged in the late 90s
, and indeed it is able to effectively remove the interpretive
overhead of repeatedly traversing the abstract syntax tree. However, these
previous works focused mostly on one particular analysis, or required to
completely rewrite the analyzer. Hence, it is worth to investigate the idea from
a modern perspective based on generative programming, especially for a general
abstract interpreter that models direct-style $\lambda$-calculus, imperative
features such as mutation, and different abstract domains. One technical
challenge is the binding-time engineering when non-determinism, fixed-point
iterations, different abstract domains, and staging are introduced at the same
time. In this paper, we present an end-to-end design and implementation of
staging an abstract interpreter; that means not only the interpreter that
traverses the abstract syntax tree, but also the data structure of abstract
domains, abstract environment and heap, and the fixed-point iteration are
staged.

\todo{rewrite this paragraph}
On the other hand, the abstract interpreter, as a semantic artifact, should be
written in a style that is easy for people to communicate the formulation and
abstraction, but also can be implemented efficiently. As the slogan of
multi-stage programming said, "abstraction without regret", we draw a connection
between high-level description and efficient implementation of abstract
interpreters, just like the connection between concrete interpreters and
compilers drawn by the Futamura projections. Particularly, we show an easy but
systematic way of adding stage annotations to the abstract interpreter without
changing the code of the interpreter skeleton, which is shared between four concrete
or abstract, unstaged or staged interpreters. We use the LMS framework for staging,
which allows us only use types to annotate the binding-time. Therefore, the
proposed approach bridges the gap between designing sound static analysis and
implementing efficient program analyzer.

%Of course, all interpreters are metalinguistic abstractions, but some
%interpreters are more "abstract" than others \todo{maybe rephrase}.
%Particularly, we also would like a systematic approach to optimize program
%analyzers and meanwhile minimally modifying the analyzer programs.

When staging a concrete interpreter, the programmers need to distinguish static
and dynamic values --- the given program to be executed by the interpreter is
classified as static because it is known at compile-time, and the inputs to that
program are dynamic. However, when staging an abstract interpreter, this
distinction does not exist anymore. Because the abstract interpreter
instantiates all the inputs as some form of abstract values, which are usually
top elements in their abstract domains and are also statically known. Then what
is the point of staging if there is no such distinction? A surprising by-product
of thinking about this question is to realize that we can apply the staged
abstract interpreter on \textit{open} programs, and the free variables
representing other parts of the program (e.g., libraries) are dynamic inputs,
therefore we obtain a modular analysis through staging, mechanically.
\note{TR: I don't understand this. The program structure is static, the abstract values
are still dynamic, no? They change in every iteration of the fixpoint algorithm}
\note{TR: why is it a big deal. On Closed programs, we obtain a constant factor,
but modular analysis has different asymptotics}

When targeting higher-order programs, either staged concrete interpreters or
staged abstract interpreters are able to compile a closure, i.e., specialize the
call of interpreter with respect to the body expression of the lambda term
without knowing the actual argument value. By generalizing this observation, we
can actually specialize the abstract interpreter with any open programs, which
unexpectedly leads to a modular analysis and improves scalability. An open
program contains free variables, which represent other parts of the program and
will be left as dynamic values. For instance, one challenge in static analysis
of modern software is that the programs are usually shipped with large library
code \cite{toman_et_al:LIPIcs:2017:7121}, for example, it has been shown that
analyzing a simple "Hello World" program in Java depends on additional 3,000
classes in the library \cite{DBLP:conf/oopsla/KulkarniMZN16}. A precise
whole-program analysis formulated as abstract interpreters can be very expensive
due to the scale of the program and the inherent complexity of the algorithm.
However, analyzing these libraries is sometimes unnecessarily repeated. When
applying the staged abstract interpreter on such open programs, we leave the
unknown arguments and calling contexts as dynamic values and generate partial
analyzing result which is represented as a residual program. The partial
analyzing result can be reused and composed with the analyzing result of
application code when available later. Therefore we mechanically obtain a
modular analysis through staging, even though the original algorithm is
formulated as a whole-program analysis.

It has been observed that partially applying context-sensitivity on selected
portions of the program could improve the precision and efficiency
\cite{zipper2018, Kastrinis:2013:HCP:2491956.2462191}. We show that staging
abstract interpreters as an approach to effectively implement hybrid
context-sensitivity\todo{}.
\fi

%\note{Expression problem \cite{DBLP:conf/icfp/GibbonsW14, DBLP:conf/ecoop/KrishnamurthiFF98, expproblem},}

>>>>

Statically analyzing semantic properties of a program is a widely known
undecidable class of problem. Abstract interpretation as a lattice-based
approach to construct sound static analysis was proposed by
\citet{DBLP:conf/popl/CousotC77}. With the Galois connection, one can
approximate to the program runtime behaviors by computing the fixed points on
the abstract domain. Despite the tremendous theoretical development of abstract
interpretation over the years, on the side of pragmatics, constructing
artifacts and analyzers that perform sound abstract interpretation was 
considered abstruse and complicated in practice for a long time.

In recent years, we observe a rich progress on how we can construct abstract
interpreters from systematic principles, instead of ad-hoc engineerings. A
notable one is the Abstracting Abstract Machine (AAM) methodology
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10}, which uncovers an
approach to derive sound abstract interpreters from their concrete counterparts,
where the soundness can be easily established by examining the transformation of
semantic artifacts. For example, a CEK machine
\cite{DBLP:conf/popl/FelleisenF87} for concrete execution can be easily
refactored to an effective 0-CFA control-flow analysis
\cite{Shivers:1988:CFA:53990.54007, Midtgaard:2012:CAF:2187671.2187672} by first
tweaking the environment dereference as a nondeterministic choice, then
allocating continuations in the environment, and constraining the addresses
space to be finite. This systematic abstraction approach can be tailored to
different analyses \todo{cite} and and has been applied to multiple variants of
small-step abstract machines \cite{DBLP:journals/jfp/HornM12,
DBLP:conf/icfp/HornM10, Sergey:2013:MAI:2491956.2491979} and big-step
definitional interpreters \cite{Wei:2018:RAA:3243631.3236800,
DBLP:journals/pacmpl/DaraisLNH17, Keidel:2018:CSP:3243631.3236767}.

More pragmatically, as an implementation strategy, several purely functional
programming approaches to build abstract interpreters were emerged, for
examples, using monads or arrows. The pure approach provides much benefits: 1)
the abstract interpretation artifacts are compositional and can be constructed
modularly, 2) one can easily prove the soundness of analysis, and 3) one can
easily reason the correctness of implementation by leveraging equational
reasoning. For example, monadic abstract interpreters \todo{two level semantics,
arrow, Galois transformers}

However, besides the intrinsic complexity of static analysis, there are
additional abstraction penalties with these high-level programming approaches.
First, similar to concrete interpreters, the abstract interpreter analyzes the
program by traversing the abstract syntax tree, which poses the interpretive
overhead, e.g., pattern matching on the AST and recursive calls on the sub
expressions. Those kind of overhead can be negligible if the abstract
interpreter only runs on the program once, but also can be accumulated
significantly if it runs repeatedly, for example, on libraries. Second, the
abstract interpreter written in pure languages usually extensively uses effect
systems to model the behaviors of abstract interpretatation, such as
nondeterminism. For example, \citet{DBLP:journals/pacmpl/DaraisLNH17} and
\citet{Sergey:2013:MAI:2491956.2491979}'s monadic abstract interpreters use
monad transformers; the very recent \citet{Githubsemantic}'s Semantic framework
uses extensible effects. The effect systems may introduce additional performance
overhead.

In this paper, we propose an abstraction-without-regret approach to eliminate
those performance penalties for abstract interpreters written in purely
functional programming style. In short, our approach borrows ideas from program
specialization and high-performance DSLs: 1) Using multi-stage programming at
type level (the Lightweight Modular Staging framework
\cite{DBLP:conf/gpce/RompfO10}, in this paper), we can
specialize the abstract interpreter with respect to the input program and then
generate efficient low-level code that does the actual analysis. 2) We use the
tagless-final approach to embed the target language interpreter in a high-level
host language (Scala, in this paper), allowing the user to implement different
semantics modularly, including the abstract semantics, without changing the main
generic interpreter. This design choice allows user to derive staged abstract
interpreters without intrusive changes to the unstaged one. Therefore, our
approach is no regret in the sense of both performance and least engineering
efforts to achieve such performance. We elaborate these two main ideas in detail
as follow.

\paragraph{Futumura Projection of Abstract Interpreters}

specialization of interpreters/abstract interpreters

\paragraph{Generic Interpreters}

specializing interpreters is not new, Futumura projections, four different tasks
for language designers and engineers, writing definitional interpreters for the
language (concrete semantics close to deno/op semantics), compilers for the
language (translate to another language, low-level), writing static analyzer for
the language (abstract semantics property we want to analyze), and a performant
static analyzer (ultimate goal of this paper). shared a single generic
interpreter

\iffalse
On the other side, static analysis is a tradeoff between performance and
precision: higher precision usually leads to longer running time.

4. Existing method to improve the performance is adhoc, engineering heavy, require to rewrite the optimized version, therefore harder to reason about the correctness
6. program analyzers are also meta-programs, they manipulate other programs as data objects
\fi
